A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-01 paper.tex
Draft of eARG section

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage[hidelinks]{hyperref}

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\end{enumerate}

@@

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
However, in parallel with this tendancy to decouple the term
Ancestral Recombination graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model an ensemble of correlated
potential ancestral trees and has been of considerable
and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
A vital distinction between the ARG (process) and ASG, however,
is that there is a hard distinction between the stochastic process
(the graph) and an observable realisation (a single tree) in the ASG
because the it encodes \emph{potential} ancestral trees.
Similarly, the Ancestral Gene Transfer Graph
models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].

@@

\subsection*{Node versus edge annotation}

% What do we mean by an EARG, informally?
\citet{griffiths1996ancestral} introduced an encoding for the ARG structure, which
is elegant and mathematically economical. In this encoding, the
fundamental units correspond directly to \emph{events}
that occur in the underlying stochastic process they were modelling.
We will refer to this ARG encoding as an ``event ARG'', or eARG:
an example is given in Fig.~\ref{fig-arg-data-structure}B.
Since each node in the graph corresponds to an event, the graph
contains both recombination nodes (node \textsf{d}) and common ancestor
nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

% What do we mean by an eARG, formally?
Formally, we can define the classical eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with that node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

% % Not sure if we actually include this, but it's good to work it through
% % to make sure the notation works. Maybe something to stick into
% % an appendix
% The most fundamental operation that we need to perform on an ARG
% data structure is extracting the local tree topology at a particular
% position along the genome. The basic strategy is to traverse
% the graph upwards, building the tree as we visit each node.
% At a recombination node, when we have a choice of two parents,
% we follow the path that corresponds to the position that we
% are building the tree for. More precisely, let $S$ be the set
% of sample nodes (usually the leaves of the graph)
% and $0 \leq x < L$ be a position along the genome, and suppose
% we are building a mapping $p_u$ that defines the parent of node
% $u$ at $x$.
% Let $N$ be a set of
% nodes we are currently constructing the tree at,
% and initialise $N \leftarrow S$.
% Then, while $|N| > 0$, we add branches to the
% tree as follows. Let $u$ be an arbitrary element of $N$ (which
% we remove).
% % This isn't right: we're not handling the case where there's
% % no edges in E where the child is $u$.
% Then, if $\sigma(u) < x$ we let $j$ be the
% smallest value such that $e_{j,1} = u$, and otherwise
% let $j$ be the largest value such that $e_{j,1} = u$.
% Then, set $v \leftarrow e_{j, 2}$,
% $p_u \leftarrow v$
% and $N \leftarrow N \cup \{v\}$.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

Another limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''.
Cite~\citep{wiuf2000coalescent}.
Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
\subsection*{Issues with event-based encoding}

Even with the change to edge annotations, a focus on specific
types of events can limit the types of ancestral history that can
be described by an eARG. As it is derived from the coalescent
with recombination, a standard eARG explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

Firstly, complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}. Thus, we can
expect datasets to span the full continuum from non-recombining
short genomes to very long genomes with high-levels of recombination
[rephrase, and some cites?],
all of which have high-quality whole genome data.

Secondly, the scale of modern datasets presents a major challenge
to the assumption that $n \ll N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size can be substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
For example,
the US dairy cattle database alone comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is the bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across
13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

For these reasons, we suggest that instead of defining the nodes in an
ARG as events, which limits us to certain models of inheritance,
we use a more flexible interpretation where they are defined as genomes.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

\newcommand{\noderef}[1]{\textsf{#1}}
% \newcommand{\msprime}[0]{\texttt{msprime}}

@@

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. ``show how some of the nodes are not required for useful interpretation
% of ancestry''.
\item Despite its status as ``the ideal data structure for
population genomic analysis''~\citep{rasmussen2014genome}, there has been
little discussion of its computational properties. We discuss the
relationship between the classical concept of an ARG and the
recently introduced succinct tree sequence data structure.
\end{enumerate}

@@

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
However, in parallel with this tendancy to decouple the term
Ancestral Recombination graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model an ensemble of correlated
potential ancestral trees and has been of considerable
and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
A vital distinction between the ARG (process) and ASG, however,
is that there is a hard distinction between the stochastic process
(the graph) and an observable realisation (a single tree) in the ASG
because the it encodes \emph{potential} ancestral trees.
Similarly, the Ancestral Gene Transfer Graph

models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].

@@

In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occured in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we call it an ``event ARG``, or eARG for brevity.
An example of this form of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an ARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This formulation is equivalent to having different ``types''
of node, and that breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
produced by programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG (the most basic downstream
operation for an ARG data structure).
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event above a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% TODO update this when then later sections have been drafted.
As we see in sections XXX and YYY there are substantial limitations
of expressivity and computational performance when using this eARG
encoding. More fundamentally, however, the general approach
of classifying nodes according to a discrete set of events
and associating genome information with those nodes is intrinsically
limiting. As we saw in the previous section, the original ARG
definitions were derived directly from the coalescent with recombination
stochastic process, and the types of ancestry that can be
described are therefore limited by the assumptions of that
model. While the coalescent as a model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.

% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$. This
assumption ensures that multiple events (i.e., simultaneous
coalescence and recombination) don't occur at the same
time, which is reflected in the eARG definitions.
This once-reasonable assumption has been overtaken by
the data we have available.
% This can be tightened up a bit, I think. We may need to drop a
% few references.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
Such datasets are also available for other species,
for example, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

Another assumption of the coalescent with recombination which prevents complex
combinations of events occuring simultaneously, is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
Clearly multiple recombinations (and other complex events) will
occur in the ancestry of such datasets, and we should not be
limited in what we can infer by the data structure that we
use to represent the results.

The eARG encoding consists of two types of event
(common ancestor and recombination) but nature is not so parsimonious
and there are many other processes that will affect the
ancestry of samples.
For example, gene conversion between homologous
chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% Some refs for where people are inferring ancestry in the
% presence of GC? Isn't this what ClonalFrame etc are all about?
Similarly, the eARG encoding cannot easily represent multiple
recombinations on a chromosome, or the transmission of multiple
chromosomes.
What we can infer about ancestral histories and
processes should not be limited by the data structure that we
use to represent such ancestry. The eARG is simply one way
to \emph{encode} such history, and, as we have shown,
one that has significant limitations of expressivity
derived from its historical origins in the idealised
coalescent with recombination model.

\section*{Node annotations}
[FIXME: this is not the intended transition or next step, just putting this
content here for now.]

@@
